Surrogate model for an aligned-spin effective one body waveform model of binary neutron star inspirals using Gaussian process regression

### 2018-11-28 paper/paper_no_event.tex
Adding version of paper without PE for GW170817.

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. 
%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz. With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

%This work improves on the surrogate developed in Ref.~\cite{LackeyBernuzziGalley2017} for a nonspinning BNS EOB waveform using the description for the point-particle and tidal terms by Bernuzzi {\it et al}.~\cite{Bernuzzi:2014owa}. Whereas in Ref.~\cite{LackeyBernuzziGalley2017} the authors constructed a time-domain surrogate with the goal of reproducing the original waveform model as accurately as possible, in this work, we construct a frequency-domain model that can be extended down to arbitrarily low frequencies and does not require an online Fourier transform for each waveform evaluation. It also enables additional techniques to accelerate likelihood evaluations such as reduced order quadrature~\cite{Antil2013, CanizaresFieldGair2013, CanizaresFieldGair2015, Smith:2016qas} and multi-banding~\cite{VinciguerraVeitchMandel2017}.

We organize the paper as follows. In Section II we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section III, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section IV. In section Section V we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss future improvements in Section VI. We also give the expressions for the TaylorF2 base model in the Appendix.

@@

%% ________________________________________________________
\section{Surrogate model}

@@

%\begin{algorithm}[H]
%\caption{Uncertainty Sampling}
%\label{alg:uc}
%\begin{algorithmic}[1]
%\State {\bf Input:} Initial surrogate 
%\For{$i = 1 \to N_{\rm New}$}
%
%\State Construct the GPR error estimator (Eq.~\eqref{eq:errorrms}). In practice this can be done by specifying the samples $\bx$, the hyperparameters $\btheta$, and dummy data for $\by$ since Eq.~\eqref{eq:errorrms} does not depend on $\by$.
%
%\State Find $({\bm x}_{\rm max}, \epsilon_{\rm rms})$ that maximizes Eq.~\eqref{eq:errorrms} over the parameter space.
%
%\State Add ${\bm x}_{\rm max}$ to the list of samples: ${\bm x} = [{\bm x}, {\bm x}_{\rm max}]$.
%
%\EndFor
%\State {\bf Output:} Updated surrogate
%\end{algorithmic}
%\end{algorithm}

@@

%% ________________________________________________________
\section{Performance of the surrogate}

@@

\subsection{Injection runs}

@@

\subsection{GW170817}

Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817. In the PE paper and catalog papers~\cite{}, the original \texttt{SEOBNRv4T} model was used with the parallelized PE code RapidPE. However, there were uncertainties about its ability to accurately recover the tidal parameter.

We use the priors discussed in~cite{O2CatalogPaper}.

%% ________________________________________________________
\section{Discussion and future work}

>>>>

Fast and accurate waveform models are necessary for measuring the properties of inspiraling binary neutron star systems such as GW170817. We present a completed version of the aligned-spin binary neutron star waveform model using the effective one body formalism known as \texttt{SEOBNRv4T}. This model includes the quadrupolar and octopolar adiabatic and dynamical tides, and we improve it by including the spin-induced quadrupole moment effect and a prescription for tapering the end of the waveform to agree with numerical simulations. The resulting model has 14 intrinsic parameters, so we use universal relations that approximate all matter effects in terms of the leading quadrupolar tidal parameters. 
%This reduces the number of intrinsic parameters to six: two masses, two aligned-spins, and two quadrupolar tidal parameters. 
Unfortunately, the current implementation can take up to an hour to evaluate, and this is too slow for many parameter estimation codes that require \red{$\mathcal{O}(10^7)$} sequential waveform evaluations. We therefore construct a fast, frequency-domain surrogate of this model using Gaussian process regression. The resulting surrogate has a maximum mismatch of \red{$4.5\times 10^{-4}$} for the Advanced LIGO detector, and requires \red{$\sim 0.13$~s} to evaluate for a waveform with a starting frequency of 20~Hz. 
%With this surrogate we do a Bayesian analysis of GW170817 and measure the combined tidal parameter $\tilde\Lambda$ to be \red{$x_{-y}^{+z}$} (median and 90\% highest posterior density interval).

@@

We organize the paper as follows. In Section~\ref{sec:eob} we provide an overview of the \texttt{SEOBNRv4T} waveform model and the approximations used to reduce the dimensionality of the parameter space. In Section~\ref{sec:surrogate}, we describe the details of building a surrogate model, choosing training set waveforms, and evaluating the final model. We then compare the accuracy and speed of the surrogate to the original model in Section~\ref{sec:performance}. In Section~\ref{sec:pe} we verify that the surrogate can correctly extract the parameters of injected waveforms and estimate the parameters of GW170817. Finally, we discuss future improvements in Section~\ref{sec:discussion} and give the expressions for the TaylorF2 base model in the Appendix.

@@

%% ________________________________________________________
\section{Surrogate model}
\label{sec:surrogate}

@@

@@

%% ________________________________________________________
\section{Performance of the surrogate}
\label{sec:performance}

@@

%\subsection{Injection runs}

@@

%\subsection{GW170817}
%
%Having confirmed that the surrogate accurately recovers the injected waveform parameters, we now do parameter estimation for GW170817. In the PE paper and catalog papers~\cite{}, the original \texttt{SEOBNRv4T} model was used with the parallelized PE code RapidPE. However, there were uncertainties about its ability to accurately recover the tidal parameter.
%
%We use the priors discussed in~cite{O2CatalogPaper}.

%% ________________________________________________________
\section{Discussion and future work}
\label{sec:discussion}
